Matching the Time of Emergence of Transboundary Fish species to Lead Time for Policy Response Under Climate Change

### 2020-10-14 Scripts/Manuscript.Rmd
First draft to share with authors

---
title: Matching the Time of Emergence of Transboundary Fish Stocks to Lead Time for Policy Response Under Climate Change
author: 'Authors: Juliano Palacios-Abrantes*^1^, Thomas Frölicher^2^, Gabriel Reygondeau^13^, U. Rashid Sumaila^1^, Colette Wabnitz^14^, and William W.L. Cheung^1^'
date: 'Affiliations: ^1^Institute for the Oceans and Fisheries, University of British Columbia, Vancouver, Canada. ^2^Climate and Environmental Physics, Physics Institute, University of Bern.'
output:
  word_document: default
  pdf_document: default
---

@@

#### Project's Library
packages <- c(
  "tidyverse" 
  # "here", # for dbem_import `here()`
  # "data.table", #dbem_import `fread()`
  # "readxl", # for reading excell files
  # "janitor", # for clearing names
  # "geosphere", # estimate distances between points `distm()`
  # "ggrepel",
  # "zoo", # for average mean
  # "parallel", # for mclapply,
  # "sf", # for mapping
  # "st", # for mapping
  # "rgdal", #Spatial analysis
  # "tools", #Spatial analysis 
  # "zeallot", # for Juanito's map
  # "gmt", # for estimating distances between points
  # "viridis"
)

@@

<!-- # Abstract -->

<!-- The delineation of Economic Exclusive Zones (EEZs) by the United Nations Convention on the Law of the Seas (UNCLOS) in the early 80s, established political boundaries across the distribution of many fish species, creating shared species between nations. Despite not implicit, UNCLOS incentivizes nations to cooperate in the management of transboundary fisheries and most management plans emplaced are based on historical distribution of catches. Climate change represents a challenge for the joint management of fish stocks. For instance, climate-driven changes in species distributions have led to conflict between nations due to changes in the proportion of captures, quota allocation, and fishery newcomers. Recent events suggests that even changes in species that contribute a relatively small proportion of the overall catch can lead to important international conflict when no legal framework is available. Thus, it is fundamental to prepare for climate change-driven impacts such as identifying the stocks that are going to change historical proportions within neighboring EEZs, the degree by which they will change, and the time when such change is going to happen. Only by adapting management plans and policy that anticipates to these effects will international fisheries treaties be climate-ready. Here, we build on previous work to identify the transboundary stocks that will see changes in their stock-share-proportion (e.g. the proportion of the stock that each EEZ has) around the world. Moreover, we determine the time of emergence of each fish stock, defined as the year in which the stock-share-proportion will change from its historical range. Our results suggest that by the mid of the 21st century EEZs will see changes in stock-share ratio between 1 and 10% with some exceeding 25%. Most of these changes will happen before 2050 urging policy to be re-formulated to account for the possible implications of these changes. By the end of the 21st century virtually all EEZs and transboundary species will have experienced changes in stock-share-proportion relative to historical distributions. The results of this research has the potential to improve climate-ready fisheries if policy is modify to respond to the consequences presented here. -->

## Introduction

Over the last century human activities have fundamentally changed the physicochemical composition of the ocean, warming the waters, increasing the pH and reducing the oxygen concentration [@IPCC:2019to]. As a consequence, many marine species have changed their distribution [@Poloczanska:2016kk] and will continue to do so in the search of the appropriate environmental conditions needed for their subsistence [@Cheung:2010dt]. Climate change is expected to impact fisheries production at different latitudes compromising our capacity to reach international sustainability goals such as the SDG 14 - life below water [@UnitedNations2018; @Cheung:2010dt; @Barange:2014kr]. Such impacts can be offset with the proper reform of current fisheries management [@Gaines:2018sg; @Cheung:2018dh]. However, such reforms will require comprehensive plans and institutions to manage species that cross international borders, also known as shared stocks [@Gaines:2018sg; @Pinsky:2018cb].

In the early 1980's the United Nations officialized the concept of Exclusive Economic Zone (EEZs) though the Law of the Sea (UCLOS) [@UnitedNations:1986tl] partitioning the distribution of several species, thus, creating what we known today as shared stocks. As defined by the Food and Agriculture Organization, shared stocks can be classified as transboundary, those that cross neighboring EEZs; straddling, that, in addition to neighboring EEZs, also visit the adjacent high seas; and highly migratory, mainly tunas and bill-fishes that cover vast oceanic regions [@Munro:2004th]. While counties are responsible for the management of stocks within their EEZs, UNCLOS incentives actions to cooperate when stocks are shared [Article 63, @UnitedNations:1986tl]. Today, an estimate of 633 transboundary fished stocks are found in different continents and across multiple habitats averaging 48.5 million tones and representing USD 78 billion in fishing revenue between 2005 and 2010 [@FishorVisa].

The management, and thus sustainability, of fisheries that target shared stocks is jeopardized as marine species shift their known distribution to cope with a changing world [@Pinsky:2018cb]. There is evidence that such shifts have already complicated shared marine resources around the globe. In the North-eastern Atlantic, evidence shows that shared stocks managed within the European Union divisions have been shifting since the 1970's compromising the region's sustainability [@Baudron:2020dj]. In the North-eastern Pacific, sockeye salmon (*Oncorhynchus nerka*) shifted in the 1980's towards Alaska, creating a ten years conflict between Canada and the United States [@Song:2017va]. Further south, the southern stock of Peruvean anchoveta (*Engraulis ringens*) shared by Chile and Peru has gradually shifted towards Chilean waters adding to the over-exploitation of the resource [@Cashion:2018cg]. Global shifts in the distribution of tunas [@ErauskinExtramiana:2019fw] have brought additional management challenges in both the Western Indian ocean and South-western Pacific [@Miller:2007kx]. Moreover, as species continue shifting with climate change, current policies might become obsolete, and thus, require new management strategies that create equitable systems across all dependent fishing nations [@Bell:2013ga; @Pinsky:2018cb]. The objective of the current paper is to explore two concepts that will be essential in addressing the challenges that climate change will bring to the management of transboundary fisheries [@Link:2010ei].

The first concept alludes to the game theory concept of thread point, known as the minimum payoff a player is willing to receive in order to cooperate with others. If a player's future payoff drops below the thread point, such player will likely break off any collaboration agreement and proceed to act solo [@OCSumaila]. A thread point can be translated to the proportion of a species distribution that each neighboring nation has (or the quota allocation that each one gets) at the time of signing off an agreement [@OCPalacios]. Many shared fisheries operate this way having quota allocated based on some historical level [@Baudron:2020dj; @GoMA; @NOAA]. If not addressed, future changes in the thread point (hereafter called Stock Share Ratio -SSR) of jointly managed fisheries can have adverse consequences for shared fisheries [@OCPalacios; @OCSumaila]. The second concept refers to the Time of Emergence (ToE), defined as the moment when a future trend of a particular variable exceeds the historical variability [@IPCC:2019to]. Such concept has been used to look at the emergence of diverse oceanic variables [@Roggers:2015jq; @Keller:2014jv; @Frolicher:2016dg] and ecological consequences of marine heatwaves [@Cheung:2020ja]. Similarly, climate change is expected to drive species further away from their historic range limits. Assessing the ToE of shared stocks is as important as the change itself, as it could reduce potential international conflicts [@Spijkers:2017ij] and aid the strengthen of marine governance [@Pinsky:2018cb;  @Pinsky:2014by; @Link:2010ei; @Fogarty:2017eo].

Here we estimate the changes in Stock Share Ratio and the Time of Emergence of 633 transboundary stocks shared by 280 EEZs of 198 coastal countries under climate change (See methods). We rely on a species distribution and Earth system model with ten ensemble members to project the distribution of the 663 selected fish stocks under a high greenhouse gas emission scenario. We treated each species in an EEZ as a single stock due to a lack of more detailed spatial data [@FishForVisa] and only considered shared species between neighboring EEZs, rather than the species’ extended distribution (e.g. we did not include the high seas). Our results suggest that virtually all EEZs will have a change in SSR by 2030 and that 87% of them already had at least one emerging species since 2006, with significant regional and environmental differences. Towards the end of the $21^{st}$ century, all of the 280 EEZs will have at least one emerging stock and all 633 stocks would have emerged at least once. Our results have direct implications for marine governance of shared marine resources under a changing world.

## Materials and methods

### Databases and species selection

We included in this analysis 633 marine exploited transboundary species representing 80% of the catch taken from the world's EEZs between 2005 and 2014 [@FishForVisa]. Each species in within an EEZ was considered a single stock unit due to a lack of more detailed spatial information [@Teh:2015gd; @FishForVisa]. We defined the boundaries of the world EEZs (e.g., Mexico Pacific EEZ and Mexico Atlantic EEZ) using the *Sea Around Us* shapefile (updated 1 July 2015, available from http://www.seaaroundus.org) noting that it subdivides the EEZs of 198 coastal states into 280 regions (e.g., Mexico's EEZ is divided in Mexico Pacific and Mexico Atlantic) including islands territories, and determined the intersections between polygons using the R package *sf* (**Supplements Fig 1 (SuppM)**) [@PackagesfSimpleF:2018vp]. For each species and EEZ we used the Sea Around Us data to estimate the catch and fishing revenue (USD 2019 from fishing activities within the global EEZs reported as the average from the last decade (2005-2014) of available spatial data [@Zeller2016; @Tai2017; @Sumaila:2015uc]. Moreover, we classified catch trends as increasing, constant and decreasing, following Palacios-Abrantes *et al.* (*hopefully in press*). Finally, each EEZ was attributed to a geopolitical region according to the UN and biome [@Reygondeau:2019uh], and species were attributed a habitat preference [@FishBase].

### Projecting species distributions under climate change

@@

We coupled the DBEM with the Geophysical Fluid Dynamics Laboratory Earth system model (GFDL-ESM2M) to project species distribution from 1951 to 2100 under climate change [@Dunne:2012bo; @Dunne:2013db]. We run the DBEM simulating biomass and MCP from 1950 to 2100 under a “worst-case” climate change scenario according to the Representative Concentration Pathway 8.5 (RCP 8.5) used by the Intergovernmental Panel on Climate Change (IPCC) [@IPCC:2014bz; Riahi:2011dk]. According to the GFDL-ESM2M, under the RCP 8.5 scenario atmospheric temperature will increase by $3.2^{\circ}C$ by 2081-2100, relative to preindustrial levels. Because the main approach of this paper relies on understanding the spatial and temporal variation of a species distribution, we have to understand both the historical as well as the future climate-driven distribution variability, in order to infer the differences. We do this by following an ensemble member approach where each member constitutes a different initial Earth system condition [@Rodggers:2015jq; @Frolicher:2009ge]. Thus, for our experiment, in each of the ten GFDL-ESM2M-ensembles the Earth system initial condition is infinitesimally different resulting in a unique atmosphere and ocean state at each point in time. This results in a different state of internal variability for each ensemble representing a plausible future under RCP 8.5 [@Rodggers:2015jq]. Following this approach allows us to estimate a secular trend for the natural variability (e.g. background noise). At the same time, by having multiple initial-condition ensemble simulations of the GFDL-ESM2M we were able to remove the natural variability of a stock's distribution by averaging the secular trend over all ensemble members (See Estimating the Time of Emergence).

### Changes in Stocks Share Ratio

We adopted the concept of threat point as the minimum proportion of the stocks’ joint distribution between neighboring EEZs that each country will accept if they were to cooperate in the management of the stock [@OCPalacios]. For each stock and time step (1951-2100) we first aggregate the number of $0.5^{\circ}$ x $0.5^{\circ}$ grid cells where the species was present within neighboring EEZs, and then calculated the proportion that each EEZ had [@FishForVisa]. We then average the results in three time periods in order to reduce climate variability within the projections. The first time period ($t_h$) comprises (1951 - 2005) and is assumed as the historic distribution of each species, thus, representing each neighboring nation's threat point. This time frame matches the historic climate data that the GFDL uses to project climate change [@Dunne:2012bo; @Dunne:2013db]. We then selected two future periods; early $21^{st}$ century as the average of 2021 to 2040 ($t_e$), and mid $21^{st}$ century as the average of 2041 to 2060 ($t_m$). We choose these time periods to match the challenges of achieving fisheries-related UN-SDGs [@UnitedNations2018] such as SDG 14.4 (end overfishing) and SDG 1.2 (poverty reduction), aimed to be achieved by 2020 and 2030, respectively [@Singh:2017ds]. The analysis was replicated for each of the ten ensemble members capturing the variation by averaging results among ensemble members ($\mu \pm \sigma$). We then estimate the percentage change in the SSR ($\Delta{SSR}$) of each future time period ($SSR_{f}$) relative to the historic time period ($SSR_h$) as follows,

### Estimating the Time of Emergence

Knowing the time (e.g. year) in which a shared stock is going to diverge from a historical variability is key for preparing to shifting stocks [@Link:2010ei]. We employed the DBEM and the ten GFDL ensemble members described above, and developed an index of transboundary range shift, fisheries and climate change (*TI*). The index is based on the variation of a species distribution centroid in relation to the centroid of the neighboring EEZs it occupy. Each species centroid was determined by the average latitude and longitude of the grid cells containing the top 10% abundance percentile of each year derived from the DBEM. The EEZ centroid was estimated using the *st* package of *R* statistical software (**Supplements Fig 1 (SuppM)**). For each ensemble member, neighboring EEZ and species, we computed the distance between centroids assuming a perfect sphere and ignoring elevation differences using the *geosphere* package of *R* statistical software;

$$D_{ens} = acos(sin(lat_{eez})*sin(lat_{spp}) + cos(lat_{eez})*cos(lat_{spp})*cos(lon_{eez}-lon_{spp}))$$
where $lat_{eez}$ and $lat_{spp}$ are the latitudes of the EEZ and specie centroid, respectively and $lon_{eez}$ and $lon_{spp}$ are the respectively longitudes.

@@

where $D_A$ and $D_B$ represent the distance between EEZ and distribution centroids of a species in EEZs A and B, respectively for each future time step from 2006 to 2100 (*t*); *sd* is the standard deviation of the historical (*th*, 1951 - 2005) centroid distribution for $D_A$ and $D_B$. The index was then smoothed to a 10 years average. Finally, to determine the time of emergence we estimated a signal to noise ratio (SNR) and set an arbitrary threshold as follows [@Rodggers:2015jq]:

$$SNR = TREND/NOISE$$

where *TREND* is the yearly mean *TI* ($\mu$) of all ten ensemble members and *NOISE* is the standard deviation of the TREND ($\sigma$) between 1951 and 2005. Note that in some cases not all ensemble members show the presence of a species in a certain EEZ throughout the study's time frame. We presented the results only for those cases where all 10 ensemble members had results for each year between 1951 and 2100 (n = 150). We determine a $\mu \pm \sigma$ (confidence interval of 68%) to show the first year when the "signal" of the transboundary index overpasses the historic "noise". Since this was an arbitrary decision, we have included a more conservative estimation ($\mu \pm 2*\sigma$ for a 95% confidence interval). The SNR for both bottom and surface sea temperature was also estimated for statistical analysis following the SNR method.

@@

We created a vulnerability index of fisheries management to shifts in transboundary species. The index is based on two factors; the importance of the shifting species to a country's total catch and revenue from transboundary species, and the time of emergence;

$$ index' = (P_c + P_v)/ToE $$
where *P_c* and *P_v* are the proportion that the shifting species' will have in the total catch and value generated from transboundary fisheries within a county's EEZs. Here, total catch and value are represented as the average of 2005 to 2014 according to the *Sea Around Us*. The index was then normalized to a 0 to 1 scale going from less vulnerable to more vulnerable:

@@

The premise of the index is that, from a management perspective, countries whose shifting transboundary stocks provide larger societal benefits and have an early ToE will be more vulnerable to shifts in SSR than their counterpart. Similarly, the index is capable to capture different sources of societal benefits (e.g. food security trough catch or livelihood trough value). We assume that the management of fisheries that provide large societal benefits will have more pressure from stakeholders while earlier times of emergence will result in less time to act, thus resulting in higher vulnerability.

### Caveats and uncertainties

We have identified three main sources of uncertainty in our analysis and point out ways to move forward in order to produce more reliable results to inform policy dealing with international fisheries under climate change. First, in this analysis we relay in one Earth System Model (GFDL-ESM2M) and climate change scenario (RCP 8.5). While the use of multiple ESMs would allow to capture the structural uncertainty of the models [@Frolicher:2016dg], using the GFDL-ESM2M results in conservative estimates as its projections are the less extreme from the CIMIP5 models [@Bopp:2013fg]. Addressing the uncertainty in future projections is key to advance marine governance for species on the move [@Pinsky:2018cb], thus, future efforts should look at the ensemble members of multiple ESM to capture the uncertainty embedded in the models. Secondly, we set an arbitrary threshold of $\pm$ 1 s.d. (67% confidence) to determine when a SSR and ToE would happen (see methods). We tested the sensitivity of our results to a larger threshold (e.g. $\pm$ 2 s.d for a 95% confidence). While the selected threshold is more relaxed than a 2 s.d., the distribution of results do not dramatically change between thresholds with a difference in ToE of about 4 years suggesting that this is a robust threshold (Fig. S4). Thirdly, in here we present the result based on 633 species that were previously identified as shared by multiple neighboring nations [@FishForVisa]. Thus, our analysis is carried at the species level, rather than a more detailed organizational level (e.g. fish stock) due to limited spatial-specific data on species sub populations (stocks). Reproducing our analysis with spatially-explicit stock data would allow to produce more certain results in changes in both SSR and ToE of transboundary stocks, and potentially identity different types of shared stock shifts [@Link:2010ei]. However, many regions of the world lack these detailed data. Moreover, in many EEZs of the world fisheries are often managed at the species level [@FishForVisa].

### Statistical analysis

We tested the results for normality (e.g. skewness, kurtosis) and performed a couple of non-parametric Kruskal–Wallis test by ranks [@Hollander:2013wz] to investigate geopolitical and ecological differences in the ToE of transboundary stocks. Moreover, we implemented a multiple linear regression analysis to a series to explain the ToE of each species in each EEZ. Specifically, we looked at habitat preferences, catch trends, geopolitical and biome location and ToE of both bottom and sea surface temperature (See *Databases and species selection*). We hypothesize that all of these factors would have significant influence in the ToE of transboundary stocks. All analyses were run using the statistical software *R version 3.5.2 (2018-12-20; Eggshell Igloo)* with the packages *data.table* [@Packagedatatable:2019uh], *janitor* [@PackagejanitorSim:2018ur], *rfishbase* [@PackagerfishbaseR:2019th], *sf* [@PackagesfSimpleF:2018vp], *sp* [@PackagespClasses:2019tn], *tidiverse* [@PackagetidyverseE:2017vq], *tidytext* [@PackagetidytextTe:2019vk], *gmt* [@gmt] and *zoo* [@PackagezooSInfr:2019tm], *readxl* [@REF], *ggrepel* [@REF], *zeallot* [@REF], *viridis* [@REF], *cowplot* [@REF], *moments* [@REF] and *pgirmess* [@REF]. All code is available at https://github.com/jepa/TransEmergence

## Results

@@

### Changes in Stock Share Ratio (SSR)

>>>>

---
title: Matching the Time of Emergence of Transboundary Fish species to Lead Time for Policy Response Under Climate Change
subtitle: Challenges to transboundary fisheries under climate change
author: 'Authors: Juliano Palacios-Abrantes*^1^, Thomas Frölicher^2^, Gabriel Reygondeau^13^, U. Rashid Sumaila^1^, Colette Wabnitz^14^, and William W.L. Cheung^1^'
date: 'Affiliations: ^1^Institute for the Oceans and Fisheries, University of British Columbia, Vancouver, Canada. ^2^Climate and Environmental Physics, Physics Institute, University of Bern.'
output:
  word_document: 
    reference_docx: ../References/template.docx
  pdf_document: default
# bibliography: ../References/FishForVisa_Ref.bib
csl: ../References/science.csl
---

<!-- # Instructions  -->

<!-- Research articles should present a major advance and can be short or long (maximum 15,000 words) format. Submissions must include an abstract (up to 150 words), an introduction, and sections with brief informative subheadings. You may include up to ten figures and/or tables and about 60 references. Supplementary materials should be limited to information that is not essential for the general understanding of the research presented in the main text and can include data sets, figures, tables, videos, or audio files.  You may include up to a total of ten figures and/or tables (combined) throughout the supplemental text.  -->

@@

#### Project's Library
packages <- c(
  "tidyverse" ,
  "png",
  "grid"
  # "here", # for dbem_import `here()`
  # "data.table", #dbem_import `fread()`
  # "readxl", # for reading excell files
  # "janitor", # for clearing names
  # "geosphere", # estimate distances between points `distm()`
  # "ggrepel",
  # "zoo", # for average mean
  # "parallel", # for mclapply,
  # "sf", # for mapping
  # "st", # for mapping
  # "rgdal", #Spatial analysis
  # "tools", #Spatial analysis 
  # "zeallot", # for Juanito's map
  # "gmt", # for estimating distances between points
  # "viridis"
)

@@

# Abstract (150  words)

<!-- An opening sentence that states the question/problem addressed by the research AND Enough background content to give context to the study AND A brief statement of primary results AND A short concluding sentence. -->

Climate change shifts in marine species distributions is affecting the current frameworks emplaced to manage species shared fisheries. Here, we rely on a modeling framework with multiple ensemble members to identify the time frame by which shared stocks distribution will change from historical ranges, and the change in the proportion of the distrirbution that each EEZ has under climate change. Our results suggest that 60% of shared species have emerged since 2006 in 87% of the world EEZs. By 2075 all EEZs will have an emerging species and by 2097 and all species will have emerged in at least one EEZ. Average changes in the ratio of species-share between neighboring EEZs will fluctuate between 1 and 10% with some exceeding 25% by 2030 relative to today’s ratio. Our results suggest that most changes will happen before 2030 urging policy to be re-formulated to account for the possible implications of these changes.

# Introduction

Over the last century human activities have fundamentally changed the physicochemical composition of the ocean, warming the waters, decreasing the pH and reducing the oxygen concentration [@IPCC:2019to]. As a consequence, many marine species have changed their distribution [@Poloczanska:2016kk] and will continue to do so in the search of the appropriate environmental conditions needed for their subsistence [@Cheung:2010dt]. Climate change is expected to impact fisheries production worldwide compromising our capacity to reach international sustainability goals such as the SDG 14 - life below water [@UnitedNations2018; @Cheung:2010dt; @Barange:2014kr]. Such impacts can be partly offset with the proper reform of current fisheries management [@Gaines:2018sg; @Cheung:2018dh]. However, these reforms will require comprehensive plans and institutions to manage species that cross international borders, also known as shared stocks [@Gaines:2018sg; @Pinsky:2018cb].

In the early 1980’s the United Nations officialized the concept of Exclusive Economic Zone (EEZs) through the Law of the Sea (UCLOS) [@UnitedNations:1986tl], partitioning the distribution of several species, thus, creating what we know today as shared stocks. As defined by the Food and Agriculture Organization, shared stocks can be classified in three non-exclusive categories. (*i*) Transboundary, those that cross neighboring EEZs; (*ii*) straddling, that, in addition to neighboring EEZs, also visit the adjacent high seas; and (*iii*) highly migratory, mainly tunas and bill-fishes that cover vast oceanic regions [@Munro:2004th]. While countries are responsible for the management of stocks within their EEZs, UNCLOS encourages actions to cooperate when stocks are shared [Article 63, @UnitedNations:1986tl]. Today, an estimate of 633 transboundary fish species are found in different continents and across multiple habitats averaging 48.5 million tones and representing USD 78 billion in fishing revenue between 2005 and 2010 (Chapter 1).

The management, and thus sustainability, of fisheries that target shared species is jeopardized as marine species shift their known distribution to cope with a changing world [@Pinsky:2018cb]. Frequently, the rules placed to manage these species are set to historical trends [@Baudron:2020dj] and do not necessarily consider the biogeography of species [@Alexa], nor the effects of climate change [@FishForVisa]. Misalignment between policies and distributional shifts have previously resulted in unsustainable harvest and international conflict [@Miller:2013iv; @Song:2017va; @Spijkers:2017ij] and are expected to continue towards the end of the $21^{st}$ century. Thus, there is a need to understand the time when climate change will affect the sharing distribution of these species as well as the intensity in order to prepare ocean governance to species on the move [@Pinsky:2018cb; @FishForVisa; @Link:2010ei].

Here we endeavor to address this gap by estimating the Time of Emergence (ToE) [@Hawkins] and Stock Share Ratio (SSR) [@OCPalacios] of 633 transboundary species worldwide. We define ToE as the time when a change  of a particular variable exceeds the historical natural internal variability [@IPCC:2019to], a concept previously used to investigate the emergence of multiple oceanic physical and biogeochemical variables [@Roggers:2015jq; @Keller:2014jv; @Frolicher:2016dg; @Cheung:2020ja]. Moreover, SSR is adopted from Game Theory as a proxy of thread point - the minimum payoff a player is willing to receive in order to cooperate with others [@OCSumaila; @Sumaila:2013].  We rely on a species distribution model driven by output from an Earth system model with ten ensemble members to project the distribution of the 663 selected shared species [@FishForVisa] by 280 EEZs of 198 coastal countries under a high greenhouse gas emission scenario RCP8.5 (See methods). We treated each species in an EEZ as a single stock due to a lack of more detailed spatial data [@FishForVisa] and only considered shared species between neighboring EEZs, rather than the species’ extended distribution (e.g. we did not include the high seas). Our results suggest that many shared species have already emerged and shifted their SSR in most neighboring EEZs posing a threat to the sustainability of shared stocks and the resilience of international treaties.

# Materials and methods

## Databases and species selection

We included in this analysis 633 marine exploited transboundary species representing 80% of the catch taken from the world's EEZs between 2005 and 2014 [@FishForVisa]. Each species within an EEZ was considered a single stock unit due to a lack of more detailed spatial information [@Teh:2015gd; @FishForVisa]. We defined the boundaries of the world EEZs (e.g., Mexico Pacific EEZ and Mexico Atlantic EEZ) using the *Sea Around Us* shapefile (updated 1 July 2015, available from http://www.seaaroundus.org) noting that it subdivides the EEZs of 198 coastal states into 280 regions (e.g., Mexico's EEZ is divided in Mexico Pacific and Mexico Atlantic; Fig. S1) including islands territories, and determined the intersections between polygons using the R package *sf* [@PackagesfSimpleF:2018vp]. Each EEZ was attributed to a geopolitical region according to the UN and biome [@Reygondeau:2019uh], and species were attributed a habitat preference [@FishBase]. For each species and EEZ we used the *Sea Around Us* data to estimate the catch and fishing revenue from fishing activities within the global EEZs. We report both catch and revenue the average from the last decade (2005-2014) of available spatial data [@Zeller2016; @Tai2017; @Sumaila:2015uc]. The monthly average US Consumer Price Index (CPI) according to the U.S. Bureau of Labor Statistic (https://www.bls.gov/cpi/) was used to convert the original 2010 real USD value to 2019 real USD.

## Projecting species distributions under climate change

@@

We coupled the DBEM with the Geophysical Fluid Dynamics Laboratory Earth system model (GFDL-ESM2M) to project species distribution from 1951 to 2100 under climate change [@Dunne:2012bo; @Dunne:2013db]. We run the DBEM simulating biomass and MCP from 1950 to 2100 under a “worst-case” climate change scenario according to the Representative Concentration Pathway 8.5 (RCP 8.5) used by the Intergovernmental Panel on Climate Change (IPCC) [@IPCC:2014bz; Riahi:2011dk]. According to the GFDL-ESM2M, under the RCP 8.5 scenario atmospheric temperature will increase by $3.2^{\circ}C$ by 2081-2100, relative to preindustrial levels. Because the main approach of this paper relies on understanding the spatial and temporal variation of a species distribution, we have to understand both the historical as well as the future climate-driven distribution variability, in order to infer the differences. We do this by following an ensemble member approach where each member constitutes a different initial Earth system condition [@Rodggers:2015jq; @Frolicher:2009ge]. Thus, for our experiment, in each of the ten GFDL-ESM2M-ensembles the Earth system initial condition is infinitesimally different resulting in a unique atmosphere and ocean state at each point in time. This results in a different state of internal variability for each ensemble representing a plausible future under RCP 8.5 [@Rodggers:2015jq]. Following this approach allows us to estimate a secular trend for the natural variability (e.g. background noise) and isolate the climate signal of a stock’s distribution by averaging the secular trend over all ensemble members.

## Estimating the Time of Emergence of transboundary species

Knowing the time (e.g., year) in which a shared stock is going to diverge from a historical variability is key for preparing to shifting species [@Link:2010ei; @FishForVisa]. Here, we coupled the DBEM with the ten GFDL ensemble members described above, and developed an index of transboundary range shift, fisheries and climate change (*TI*). The index is based on the variation of a species distribution centroid in relation to the centroid of the neighboring EEZs it occupies. Each species centroid was determined by the average latitude and longitude of the grid cells containing the top 10% abundance percentile of each year derived from the DBEM. The EEZ centroid was estimated using the *st* package of *R* statistical software (**Supplements Fig 1 (SuppM)**). For each ensemble member, neighboring EEZ and species, we computed the distance between centroids assuming a perfect sphere and ignoring elevation differences using the *geosphere* package of *R* statistical software;

$$D_{ens} = acos(sin(lat_{eez})*sin(lat_{spp}) + cos(lat_{eez})*cos(lat_{spp})*cos(lon_{eez}-lon_{spp}))$$
where $lat_{eez}$ and $lat_{spp}$ are the latitudes of the EEZ and specie centroid, respectively and $lon_{eez}$ and $lon_{spp}$ are the respective longitudes.

@@

where $D_A$ and $D_B$ represent the distance between EEZ and distribution centroids of a species in EEZs A and B, respectively for each future time step from 2006 to 2100 (*t*); *sd* is the standard deviation of the historical (*th*, 1951 - 2005) centroid distribution for $D_A$ and $D_B$. The index was then smoothed to a 10 years average.  Finally, we determined the time of emergence setting an arbitrary threshold following Rodggers (2015):

$$ToE = TREND/NOISE$$

where *TREND* is the yearly mean *TI* ($\mu$) of all ten ensemble members and *NOISE* is the standard deviation of the TREND ($\sigma$) between 1951 and 2005. 
<!-- Note that in some cases not all ensemble members show the presence of a species in a certain EEZ throughout the study's time frame. We presented the results only for those cases where all 10 ensemble members had results for each year between 1951 and 2100 (n = 150). -->
We estimated the mean and one standar deviation ($\mu \pm \sigma$ for a confidence interval of 68%) to show the first year when the "signal" of the transboundary index overpasses the historic "noise". Since this was an arbitrary decision, we have included a more conservative estimation ($\mu \pm 2*\sigma$ for a 95% confidence interval). The SNR for both bottom and surface sea temperature was also estimated for statistical analysis following the SNR method.

@@

We created a vulnerability index of fisheries management to shifts in transboundary species. The index is based on two factors; the ToE and the importance of the shifting species to a country's total catch and revenue from transboundary species;

$$ index' = (P_c + P_v)/ToE $$
where *P_c* and *P_v* are the proportion that the shifting species' will have in the total catch and value generated from transboundary fisheries within a county's EEZs, respectively. Here, total catch and value are represented as the average of 2005 to 2014 within the global EEZs according to the *Sea Around Us*. The index was then normalized to a 0 to 1 scale going from less vulnerable to more vulnerable:

@@

## Changes in transboundary Species Share Ratio

We adopted the concept of threat point as the minimum proportion of the species’ joint distribution between neighboring EEZs that each country will accept if they were to cooperate in the management of the stock [@OCPalacios]. For each species and time step (1951-2100) we first aggregate the number of $0.5^{\circ}$ x $0.5^{\circ}$ grid cells where the species was present within neighboring EEZs, and then calculated the proportion that each EEZ had [@FishForVisa]. We then average the results in three time periods in order to reduce climate variability within the projections. The first time period ($t_h$) comprises (1951 - 2005) and is assumed as the historic share distribution of each species, thus, representing each neighboring nation's threat point. This time frame matches the historic climate data that the GFDL uses to project climate change [@Dunne:2012bo; @Dunne:2013db]. We then selected two future periods; early $21^{st}$ century as the average of 2021 to 2040 ($t_e$), and mid $21^{st}$ century as the average of 2041 to 2060 ($t_m$). We choose these time periods to match the challenges of achieving fisheries-related UN-SDGs [@UnitedNations2018] such as SDG 14.4 (end overfishing) and SDG 1.2 (poverty reduction), aimed to be achieved by 2020 and 2030, respectively [@Singh:2017ds]. The analysis was replicated for each of the ten ensemble members capturing the variation by averaging results among ensemble members ($\mu \pm \sigma$). We then estimate the percentage change in the SSR ($\Delta{SSR}$) of each future time period ($SSR_{f}$) relative to the historic time period ($SSR_h$) as follows,

## Statistical analysis

We tested the results for normality (e.g. skewness, kurtosis) and performed a couple of non-parametric Kruskal–Wallis test by ranks [@Hollander:2013wz] to investigate geopolitical and ecological differences in the ToE of transboundary species. 
<!-- Moreover, we implemented a multiple linear regression analysis to a series to explain the ToE of each species in each EEZ. Specifically, we looked at habitat preferences, catch trends, geopolitical and biome location and ToE of both bottom and sea surface temperature (See *Databases and species selection*). We hypothesize that all of these factors would have significant influence in the ToE of transboundary species. -->
All analyses were run using the statistical software *R version 3.5.2 (2018-12-20; Eggshell Igloo)* with the packages *data.table* [@Packagedatatable:2019uh], *janitor* [@PackagejanitorSim:2018ur], *rfishbase* [@PackagerfishbaseR:2019th], *sf* [@PackagesfSimpleF:2018vp], *sp* [@PackagespClasses:2019tn], *tidiverse* [@PackagetidyverseE:2017vq], *tidytext* [@PackagetidytextTe:2019vk], *gmt* [@gmt] and *zoo* [@PackagezooSInfr:2019tm], *readxl* [@REF], *ggrepel* [@REF], *zeallot* [@REF], *viridis* [@REF], *cowplot* [@REF], *moments* [@REF] and *pgirmess* [@REF]. All code is available at https://github.com/jepa/TransEmergence

# Results

@@

## Time of Emergence of transbounudary species

We estimated the ToE between neighboring EEZs for 633 transboundary species, that is, the year where the future stock's distribution (2006-2100) overpasses the historical variation (1951-2005). Since 2006, 87% of the world EEZs have already seen at least one emerging stock and 60% of the transboundary species analyzed have emerged in at least a set of neighboring EEZ (e.g. *Engraulis ringens* shared by Chile and Peru emerged in 2010). species will continue to emerge steadily towards 2097 when all 633 species will have emerged in at least one neighboring EEZ (Fig. 1). The number of EEZs with emerging species will increase steeply until 2020 when it seems to flatten approaching 2081 when all of the EEZs analyzed here will have at least one emerging stock (Fig. 1).

```{r Figure1, eval = T,  echo = F, fig.width = 5, fig.height=5, fig.cap =  "Cumulative increase in time of emergence by Exclusive Economic Zone and transboundary species in absolute numbers and percentage."}

<!-- ![Figure 1. Cumulative increase in time of emergence by Exclusive Economic Zone and transboundary species in absolute numbers and percentage.](../Figures/Fig2.png) -->

Most of the world EEZ’s will see an average ToE across transboundary species before 2040 (Fig. 2). In a few EEZs species will have a relative late average ToE (e.g. Honduras average ToE is 2071) while 54 of the 280 EEZs (19%) have already seen their species emerged (e.g. average ToE across all species <= 2020). A higher number of emerging species will happen between latitudes 0$^o$ and 50$^o$N corresponding to regions of relatively small EEZs with multiple borders (e.g. The Caribbean and the Mediterranean) (Fig.2). In most cases, the current participation of the emerging species in a country's total fishing revenue is less than 50% (Fig. 2 - Polygons). However, in some cases emerging species represent 85% of current transboundary fishing revenue (e.g. Peru) or more (e.g. Marshall Islands, 92%). Only in very few cases like Brazil and Alaska's polar EEZ, no specie will emerge between 2006 and 2100.

```{r Figure2, eval = T,  echo = F, fig.width  = 12, fig.height =10, fig.cap =  "Average time of emergence of transboundary specie. Land polygon shows the vulnerability index. Exclusive Economic Zone polygon represents the average time of emergence across all species. Density plot on the right shows the latitudinal number of emerging species across EEZ centroids (10^o running mean)."}
